OR Multi Disturbance

### 2020-12-30 OR_original_model.Rmd
Flushed out analysis of basic real-world application. Methods: model description

## Load Calibration
load("fit.RData")

@@

  for(t in 1:2){

@@

      ## draw new IC
    if(run.adjustment){
    Na <- multi.ens.adj(
      Xf   = Nf,
      cf   = d.forecast,
      mu.f = rbind(priors$mufN,priors$mufD),
      Pf   = abind::abind(solve(priors$pfN),
                            solve(priors$pfD),
                            along=3),
      Xa = Xa,
      ca = ca)
    } else {
      Na = cbind(rep(0,nrow(Nf)),Nf)
    }
    Ca <- Na[,1] ## updated classf assignments
    Na <- tobit(Na[,-1])
    #  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
    #  Na <- dat[sel.ic,nsel]
    Analysis[[t]] <- list(Na = Na,
                          Ca = Ca,
                          CI = apply(dat,2,quantile,c(0.025,0.25,0.5,0.75,0.975))
    )
  }
  return(list(Forecast=Forecast,Analysis=Analysis))
}
```

@@

## analyze sites - just one disturbance
```{r}
i = 181  # 177:263 = #fire
NT = floor(ncol(rad)/365)
Yi = matrix(NA,NT,3)  ## observed data
Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed
R = diag(rep(Rw/1000,3))    ## observation error   ### div 1000 is test
Q = diag(c(Ql,Qs,Qw)^2)*10  ## process error
dist <- list(              ## disturbance parameters
  p.dist = d.prior[4],     ## disturbance probability
  mu0 = c(0.2,PBBAR[4,4]),   ## % retention [leaf,agb], first parameter made up, should include soil
  V0 =  diag(c(0.15,PBVAR[4,4])^2), #variance in retention
  alloc.soil = 0.05        ## made up  
)
test = SSSD.model(Y = Yi,R=R,Q=Q,
                  PAR=rad[i,],Xic=ICprior,THETA=theta[1:11],
                  dist = dist)
```

@@

>>>>

## Load Calibration
load("fit.RData")
Qsoil = Qs

@@

  for(t in 1:NT){

@@

      ## draw new IC
    if(run.adjustment){
    Na <- multi.ens.adj(
      Xf   = Nf,
      cf   = d.forecast,
      mu.f = rbind(priors$mufN,priors$mufD),
      Pf   = abind::abind(solve(priors$pfN),
                            solve(priors$pfD),
                            along=3),
      Xa = Xa,
      ca = ca)
    } else {
      Na = cbind(rep(0,nrow(Nf)),Nf)
    }
    Ca <- Na[,1] ## updated classf assignments
    Na <- tobit(Na[,-1])
    #  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
    #  Na <- dat[sel.ic,nsel]
    Analysis[[t]] <- list(Na = Na,
                          Ca = Ca,
                          CI = apply(dat,2,quantile,c(0.025,0.25,0.5,0.75,0.975)),
                          mu = apply(dat,2,mean,na.rm=TRUE)
    )
  }
  return(list(Forecast=Forecast,Analysis=Analysis))
}
```

@@

## analyze sites - just one disturbance
```{r}
Rmult = c(1,0.1,0.01,0.001)
Qmult = c(1,10)
Pmult = c(1)
Ri = 4 #4
Qi = 1 #2
Pi = 1
#i = 181  # 177:263 = #fire
od = c(83:176,264:356)
for(i in 83:356){
  ofile = paste0("OR_SD_",
                 formatC(i, width = 3, format = "d", flag = "0"),
                 "_P",Pi,"Q",Qi,"R",Ri,".RData")
  print(ofile)
  NT = floor(ncol(rad)/365)
  Yi = matrix(NA,NT,3)  ## observed data
  Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed
  R = diag(rep(Rw,3))*Rmult[Ri]    ## observation error
  Q = diag(c(Ql,Qsoil,Qw)^2)*Qmult[Qi]  ## process error
  dtype = which(names(d.prior) == dextract[i,"t2004"])
  if(length(dtype)==0) break
  dist <- list(              ## disturbance parameters
    p.dist = d.prior[dtype]*Pmult[Pi],     ## disturbance probability
    mu0 = c(0.2,PBBAR[4,dtype]),   ## % retention [leaf,agb], first parameter made up, should include soil
    V0 =  diag(c(0.15,PBVAR[4,dtype])^2), #variance in retention
    alloc.soil = 0.05        ## made up  
  )
  test = SSSD.model(Y = Yi,R=R,Q=Q,
                  PAR=rad[i,],Xic=ICprior,THETA=theta[1:11],
                  dist = dist)
  save(test,file=ofile)
}
```

@@

```{r}
## Detect set of experiments that have been run
runs <- dir(pattern="*.RData")
runs <- runs[grepl("^OR",runs)]
exp = sapply(strsplit(runs,"_",fixed=TRUE),function(x){
  sub(".RData","",x[4],fixed = TRUE)
})
exps <- unique(exp)

## load experiments that haven't yet been summarized
# elist <- list()   ## run once
t04 = 15            ## time of disturbance assessement (2004)
tlo = 28            ## time of last observation
for(i in seq_along(exps)){
  if(is.null(elist[[exps[i]]])){
    esel <- which(exp==exps[i])
    distrate = rep(NA,nrow(dextract))
    divergence = predist = distsize = matrix(NA,nrow(dextract),6) ## pre and post disturbance biomass pools (w/uncertainty)
    for(j in seq_along(esel)){
      ## load file
      load(runs[esel[j]])
      k = as.numeric(unlist(strsplit(runs[esel[j]],"_",fixed=TRUE))[3])

      ## extract detection rate
      distrate[k]=test$Analysis[[t04]]$mu["D"]

      ## capture of disturbance change
      predist[k,1]    = test$Analysis[[t04-1]]$mu["n[3]"]
      predist[k,2:6]  = test$Analysis[[t04-1]]$CI[,"n[3]"]
      distsize[k,1]   = test$Analysis[[t04]]$mu["n[3]"]
      distsize[k,2:6] = test$Analysis[[t04]]$CI[,"n[3]"]

      ## Filter divergence
      divergence[k,1]   = test$Analysis[[tlo]]$mu["n[3]"]
      divergence[k,2:6] = test$Analysis[[tlo]]$CI[,"n[3]"]
    } ## end loop over individual runs

    elist[[exps[i]]] = list(distrate=distrate,predist=predist,
                            distsize=distsize,divergence=divergence)

  } ## end if not processed

} ## end loop over experiments

## grab matching data
obs.pre  = Yw[,t04-1]
obs.dist = Yw[,t04]
obs.tlo  = Yw[,tlo]
obs.class= dextract[,"t2004"]/10
```

## analyses

```{r}
pdf("OR_exps_diagnostics.pdf")
trim <- function(x,lo=0.0001,hi=0.9999){
  x[x<lo] <- lo
  x[x>hi] <- hi
  return(x)
}
betatrim <- function(y){
    n.obs <- sum(!is.na(y))
    (y * (n.obs - 1) + 0.5) / n.obs
}
stats <- list()
for(i in seq_along(exps)){

  ## divergence
  div = (elist[[i]]$divergence[,1] - obs.tlo)/
    (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
  div.reg    = lm(div~obs.tlo)
  div.median = median(div,na.rm = TRUE)
  div.cov    = sum(div < 1 & div > -1,na.rm=TRUE)/sum(!is.na(div))

  hist(div,main=exps[i])
  abline(v=c(-1,1),lty=2)
  plot(obs.tlo,div,col=obs.class,main=exps[i])
  abline(h=c(-1,1),lty=2)
  abline(div.reg,col=2)

  ## Disturbance detection
  distrate = elist[[i]]$distrate
  hist(distrate,main=exps[i])

  ## Disturbance skill
  ord <- order(distrate)
  plot(distrate[ord],col=obs.class[ord],main=exps[i])
  dist.bar   = mean(distrate,na.rm = TRUE)
  dist.class = tapply(distrate,obs.class,mean,na.rm=TRUE)

  ## vs predisturbance biomass
  distrate2 = betatrim(distrate)
  pred.seq = data.frame(obs.pre=seq(0,max(obs.pre,na.rm = TRUE),length=500),
                        distprop = seq(-1,1,length=500))
  dist.reg.pre  = lm(distrate ~ obs.pre)
  dist.beta.pre = betareg::betareg(distrate2 ~ obs.pre,
                                   na.action=na.omit,link="logit")
  plot(obs.pre,distrate,col=obs.class,main=exps[i])
  abline(dist.reg.pre,col=2)
  lines(pred.seq$obs.pre,predict(dist.beta.pre,newdata=pred.seq),col="green")

  ## vs proportional disturbance size
  distprop = 1 - elist[[i]]$distsize[,1]/elist[[i]]$predist[,1] ## predicted
  distprop = 1 - obs.dist/obs.pre ## observed
  dist.beta.prop = betareg::betareg(distrate2 ~ distprop,
                                   na.action=na.omit,link="logit")
  plot(distprop,distrate,col=obs.class,main=exps[i])
  lines(pred.seq$distprop,predict(dist.beta.prop,newdata=pred.seq),col="green")

  ## vs ABSOLUTE disturbance size
  distabs = obs.pre-obs.dist
  pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
  dist.beta.abs = betareg::betareg(distrate2 ~ distabs | distabs,
                                   na.action=na.omit,
                                   link="logit",
                                   link.phi = "identity",
                                   start=c(0,5,0.5,0),
                                   type="BR")
  dist.qb.abs = glm(distrate ~ distabs,
                    na.action=na.omit,
                    family=quasibinomial("logit")
  )
  dist.bin.abs = glm(round(distrate) ~ distabs,
                    na.action=na.omit,
                    family=binomial("logit")
  )
  plot(distabs,distrate,col=obs.class,main=exps[i])
  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile"),col="green")
  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile",at=0.75),col="green",lty=2)
  lines(pred.seq$distabs,predict(dist.beta.abs,newdata=pred.seq,type="quantile",at=0.25),col="green",lty=2)
  lines(pred.seq$distabs,predict(dist.qb.abs,newdata=pred.seq,type="response"),col="purple")
  lines(pred.seq$distabs,predict(dist.bin.abs,newdata=pred.seq,type="response"),col="blue")

  ## vs divergence
  absdiv = abs(div)
  dist.beta.absdiv = betareg::betareg(distrate2 ~ absdiv,
                                   na.action=na.omit,link="logit")
  dist.bin.absdiv = glm(round(distrate) ~ absdiv,
                                   na.action=na.omit,
                        family = binomial("logit"))
  pred.seq$absdiv = seq(0,max(absdiv,na.rm=TRUE),length=500)
  plot(absdiv,distrate,col=obs.class,main=exps[i])
  abline(v=c(-1,1),lty=2,col=3)
  abline(lm(distrate ~ absdiv),col=2)
  lines(pred.seq$absdiv,predict(dist.beta.absdiv,newdata=pred.seq),col="green")
  lines(pred.seq$absdiv,predict(dist.bin.absdiv,newdata=pred.seq,type="response"),col="blue")

 dist.qb.combo = glm(distrate2 ~ obs.pre + distprop +
                        distabs + absdiv + obs.class,
                      na.action=na.omit,
                      family=quasibinomial("logit"))
  dist.bin.combo = glm(round(distrate) ~ obs.pre + distprop +
                        distabs + absdiv + as.factor(obs.class),
                      na.action=na.omit,
                      family=binomial("logit"))

# dist.beta.combo = betareg::betareg(distrate2 ~ obs.pre + obs.class + distabs + absdiv,
#                                   na.action=na.omit,link="logit")

  stats[[i]] <- list(
    div.reg = div.reg,
    div.median = div.median,
    div.cov = div.cov,
    dist.bar = dist.bar,
    dist.class = dist.class,
    dist.pre =  dist.beta.pre,
    dist.prop = dist.beta.prop,
    dist.beta.abs =  dist.beta.abs,
    dist.bin.abs  = dist.bin.abs,
    dist.absdiv   = dist.beta.absdiv,
    dist.combo = dist.bin.combo
  )

}
dev.off()
```

factors affecting detection of disturbance
```{r}
Ps = as.numeric(substr(names(elist),2,2))
Qs = as.numeric(substr(names(elist),4,4))
Rs = as.numeric(substr(names(elist),6,6))

## P

## Q

## R

## Divergence
div.cov = sapply(stats,"[[","div.cov")
plot(Rmult[Rs],div.cov,col=Qs,log="x",xlab="Variance Multiplier",ylab="Filter Coverage")

div.median = sapply(stats,"[[","div.median")
plot(Rmult[Rs],div.median,col=Qs,log="x",xlab="Variance Multiplier",ylab="Median Divergance")

div.slope = sapply(stats,function(x){coef(x$div.reg)[2]})
plot(Rmult[Rs],div.slope,col=Qs,log="x",xlab="Variance Multiplier",ylab="Divergance Slope")
## Disturbance @ default
summary(stats[[4]]$dist.combo)
## nothing matters

## Disturbance @ P1Q2R2 (intermediate)
summary(stats[[4]]$dist.combo)
## detection varies with divergence & disturbance type

## Disturbance magnitude: P1Q1R4
i = 3
summary(stats[[i]]$dist.combo)  ## absolute magnitude of divergence is best predictor
summary(stats[[i]]$dist.bin.abs)
beta = coef(stats[[i]]$dist.bin.abs)
## threshold mx+b = 0
thresh = -beta[1]/beta[2]

## Disturbance type: P1Q1R4 (reduced divergence)
stats[[i]]$dist.class
tapply(distabs,obs.class,summary) ## obs dist size by class
thresh.prop = tapply(distabs,obs.class,function(x){sum(x>thresh,na.rm = TRUE)/sum(!is.na(x))}) ## prop > detection threshold
thresh.prop
stats[[i]]$dist.class
dclass = stats[[i]]$dist.class
#dclass[1] = 0
dcol = as.numeric(names(stats[[i]]$dist.class)) + 1
plot(thresh.prop,dclass,col=dcol,pch=19)
abline(0,1,lty=2,col=2)
legend("topleft",legend = dcol-1,col=dcol,pch=19)

## additional experiments to run (P1Q1R4)
```

### 2020-12-30 methods.Rmd
Flushed out analysis of basic real-world application. Methods: model description

>>>>

## VSEM (Very Simple Ecosystem Model)

To illustrate the problem of assimilating disturbance, and present potential solutions, we make use of the Very Simple Ecosystem Model (VSEM), a daily-timestep point-scale model developed by David Cameron that is available in the BayesianTools R package [@BayesianTools]. VSEM contains three carbon pools, leaf (Cl), wood (Cw), and soil (Cs), and calculates leaf area index based on specific leaf area (SLA), $LAI = SLA*Cl$. Gross Primary Productivity (GPP) is estimated based on incoming photosynthetically active radiation (PAR), light use efficiency (LUE), and Beer's Law light interception ($1-exp(-K_{ext}\cdot LAI)$) with a extinction coefficient $K_{ext}$. Autotrophic respiration, Ra, is a fixed fraction, $\Gamma$ of GPP. Of the remaining carbon (net primary productivity), a fixed fraction, $Av$, is allocated to leaves and the remainder is allocated to wood. Leaves and wood turnover into soil C based on fixed longevities, $\tau_l$ and $\tau_w$, while heterotrophic respiration, Rh, occurs based on a fixed residence time $\tau_s$. Net Ecosystem Exchange follows the atmospheric sign convention $Ra + Rh - GPP$. The VSEM was selected for its speed and ease of understanding, but its fundamental structure is analogous to much more complex, large-scale terrestrial carbon models.

## Ensemble Adjustment Kalman Filter

To
